Plant Trait Imputation

### 2024-06-11 code/L1/L1_P5_TropicalAndes_plant_imputetraits.Rmd
add visualization packages and code

# Load required packages
```{r}
library(tidyr)
library(BIEN)
library(GIFT)
library(purrr)
library(mice)
library(dplyr)
```

@@

      # Rename the mean column to TraitValue
      summary_data <- summary_data %>%
        rename(trait_value = trait_value_mean)

@@

```{r}
family_level_trait_data <- list(PlantLifespan_years_families_BIEN_df_average, FruitType_families_BIEN_df_average, GrowthForm_families_BIEN_df_average, FruitLength_mm_families_BIEN_df_average, SeedMass_g_families_BIEN_df_average, PlantHeight_m_genus_BIEN_df_average)

# Function to select and convert columns to character
select_and_convert <- function(df) {
  df %>%
    select(scrubbed_family, TraitName, TraitValue) %>%
    mutate(across(everything(), as.character))
}

# Apply the function to each dataframe in the list and combine them
all_family_traits <- family_level_trait_data %>%
  purrr::map_df(select_and_convert)
```

@@

#### Combine all genus level trait data
```{r}
## add column of trait_name for numeric traits
Dispersal.Syndrome_genus_BIEN_df$trait_name <- "Dispersal.Syndrome"
Fruit.Length_genus_BIEN_df$trait_name <- "Fruit.Length_mm"
Plant.Height_genus_BIEN_df$trait_name <- "Plant.Height_m"
Fruit.Type_genus_BIEN_df$trait_name <- "Fruit.Type"
Growth.Form_genus_BIEN_df$trait_name <- "Growth.Form"

## rename TraitValue column so it's consistent with other traits
Fruit.Type_genus_BIEN_df$TraitValue <- Fruit.Type_genus_BIEN_df$trait_value_mode_trait
Growth.Form_genus_BIEN_df$TraitValue <- Growth.Form_genus_BIEN_df$trait_value_mode_trait

## convert numeric trait values into character
Plant.Height_genus_BIEN_df$TraitValue <- as.character(Plant.Height_genus_BIEN_df$TraitValue)
Fruit.Length_genus_BIEN_df$TraitValue <- as.character(Fruit.Length_genus_BIEN_df$TraitValue)

all_genus_traits <- bind_rows(
  select(Dispersal.Syndrome_genus_BIEN_df, scrubbed_genus, trait_name, TraitValue),
  select(Fruit.Length_genus_BIEN_df, scrubbed_genus, trait_name, TraitValue),
  select(Plant.Height_genus_BIEN_df, scrubbed_genus, trait_name, TraitValue),
  select(Fruit.Type_genus_BIEN_df, scrubbed_genus, trait_name, TraitValue),
  select(Growth.Form_genus_BIEN_df, scrubbed_genus, trait_name, TraitValue)
  )

@@

rename traits to match
```{r}
# Replace trait names to match
long_plant_traits_tax <- long_plant_traits_tax %>%
  mutate(TraitName = ifelse(TraitName == "Plant.Height", "Plant.Height_m", TraitName)) %>%
  mutate(TraitName = ifelse(TraitName == "Fruit.Length", "Fruit.Length_mm", TraitName))
unique(long_plant_traits_tax$TraitName)

na_traits_family.genus <- na_traits_family.genus %>%
  mutate(TraitName = ifelse(TraitName == "Plant.Height", "Plant.Height_m", TraitName)) %>%
  mutate(TraitName = ifelse(TraitName == "Fruit.Length", "Fruit.Length_mm", TraitName))
unique(na_traits_family.genus$TraitName)
```

@@

```{r}
colnames(all_family_traits)[which(colnames(all_family_traits) == "scrubbed_family")] <- "family"
colnames(all_family_traits)[which(colnames(all_family_traits) == "trait_name")] <- "TraitName"
colnames(all_family_traits)[which(colnames(all_family_traits) == "trait_value")] <- "TraitValue"

colnames(all_genus_traits)[which(colnames(all_genus_traits) == "scrubbed_genus")] <- "genus"
colnames(all_genus_traits)[which(colnames(all_genus_traits) == "trait_name")] <- "TraitName"
colnames(all_genus_traits)[which(colnames(all_genus_traits) == "trait_value")] <- "TraitValue"
```

@@

```{r}
na_traits_after_genus <- filled_na_traits_family.genus %>%
  filter(is.na(TraitValue))
filled_traits_with_genus <- filled_na_traits_family.genus %>%
  filter(!is.na(TraitValue))
```

@@

```{r}
filled_traits_with_family <- filled_na_traits_family %>%
  filter(!is.na(TraitValue))
na_traits_after_family <- filled_na_traits_family %>%
  filter(is.na(TraitValue))
```

@@

@@

## Convert long dataframe to wide
```{r}
#remove unused columns
all_traits_with_NAs_long <- all_traits_with_NAs %>%
  select(-TraitLevel, -family, -genus)
wide_traits <- pivot_wider(all_traits_with_NAs_long, names_from = TraitName, values_from = TraitValue)
```
make NA if:
- Growth.Form is numeric
- Dispersal.Syndrome is numeric or contains multiple values
- Plant.Height = 0
- Fruit.Type is numeric
```{r}
library(stringr)
wide_traits$Growth.Form <- ifelse(is.numeric(wide_traits$Growth.Form), NA, wide_traits$Growth.Form)
wide_traits$Dispersal.Syndrome[stringr::str_detect(wide_traits$Dispersal.Syndrome, ",")] <- NA

# Convert Plant.Height_m to numeric 
wide_traits$Plant.Height_m[wide_traits$Plant.Height_m == "0"] <- NA

wide_traits$Fruit.Type <- ifelse(is.numeric(wide_traits$Fruit.Type), NA, wide_traits$Fruit.Type)
```

```{r}
wide_traits <- wide_traits %>%
  mutate_all(~sapply(., paste, collapse = ", "))

# Replace values with multiple values with NA
wide_traits <- wide_traits %>%
  mutate(Dispersal.Syndrome = if_else(str_count(Dispersal.Syndrome, ", ") > 0, NA_character_, Dispersal.Syndrome), 
         Growth.Form = if_else(str_count(Growth.Form, ", ") > 0, NA_character_, Growth.Form),
         Plant.Height_m = if_else(str_count(Plant.Height_m, ", ") > 0, NA_character_, Plant.Height_m), 
         Fruit.Length_mm = if_else(str_count(Fruit.Length_mm, ", ") > 0, NA_character_, Fruit.Length_mm), 
         Fruit.Type = if_else(str_count(Fruit.Type, ", ") > 0, NA_character_, Fruit.Type))
```

```{r}
wide_traits$Plant.Height_m <- unlist(wide_traits$Plant.Height_m)
wide_traits$Growth.Form <- unlist(wide_traits$Growth.Form)
wide_traits$Dispersal.Syndrome <- unlist(wide_traits$Dispersal.Syndrome)
wide_traits$Fruit.Length_mm <- unlist(wide_traits$Fruit.Length_mm)
wide_traits$Fruit.Type <- unlist(wide_traits$Fruit.Type)

#convert column types from list to character & numeric
wide_traits$Plant.Height_m <- as.numeric(wide_traits$Plant.Height_m)
wide_traits$Growth.Form <- as.character(wide_traits$Growth.Form)
wide_traits$Dispersal.Syndrome <- as.character(wide_traits$Dispersal.Syndrome)
wide_traits$Fruit.Length_mm <- as.numeric(wide_traits$Fruit.Length_mm)
wide_traits$Fruit.Type <- as.character(wide_traits$Fruit.Type)
```

@@

# show the missing data pattern
```{r}
md.pattern(wide_traits)
```
# Perform the imputation
```{r}
imp_model <- mice(wide_traits, method = "cart", maxit = 20)
```

@@

remove dispersal syndrome because I need to fix it 
```{r}
imputed_data <- imputed_data %>%
  select(-Dispersal.Syndrome)
```

```{r}
imputed_data %>%
  summarise_at(vars(Growth.Form:Fruit.Type), ~ sum(is.na(.)))
```

>>>>

# Load required packages
```{r}
library(tidyr)
library(BIEN)
library(GIFT)
library(purrr)
library(mice)
library(dplyr)
library(funbiogeo)
library(visdat)
library(ggplot2)
library(forcats)
```

@@

@@

```{r}
family_level_trait_data <- list(PlantLifespan_years_families_BIEN_df_average, 
                                FruitType_families_BIEN_df_average, 
                                GrowthForm_families_BIEN_df_average, 
                                FruitLength_mm_families_BIEN_df_average, 
                                SeedMass_g_families_BIEN_df_average, 
                                PlantHeight_m_families_BIEN_df_average)

# Function to select and convert columns to character
select_and_convert <- function(df) {
  df %>%
    dplyr::select(scrubbed_family, TraitName, TraitValue) %>%
    mutate(across(everything(), as.character))
}

# Apply the function to each dataframe in the list and combine them
all_family_traits <- family_level_trait_data %>%
  map_df(select_and_convert) %>%
  bind_rows()
```

@@

#### Combine all genus level trait data
```{r}
colnames(PlantLifespan_years_genus_BIEN_df_average)
colnames(FruitType_genus_BIEN_df_average)
colnames(GrowthForm_genus_BIEN_df_average)
colnames(FruitLength_mm_genus_BIEN_df_average)
colnames(SeedMass_g_genus_BIEN_df_average)
colnames(PlantHeight_m_genus_BIEN_df_average)
```

```{r}
genus_level_trait_data <- list(PlantLifespan_years_genus_BIEN_df_average, 
                                FruitType_genus_BIEN_df_average, 
                                GrowthForm_genus_BIEN_df_average, 
                                FruitLength_mm_genus_BIEN_df_average, 
                                SeedMass_g_genus_BIEN_df_average, 
                                PlantHeight_m_genus_BIEN_df_average)

# Function to select and convert columns to character
select_and_convert <- function(df) {
  df %>%
    dplyr::select(scrubbed_genus, TraitName, TraitValue) %>%
    mutate(across(everything(), as.character))
}

# Apply the function to each dataframe in the list and combine them
all_genus_traits <- genus_level_trait_data %>%
  map_df(select_and_convert) %>%
  bind_rows()

@@

@@

```{r}
colnames(all_family_traits)[which(colnames(all_family_traits) == "scrubbed_family")] <- "family"

colnames(all_genus_traits)[which(colnames(all_genus_traits) == "scrubbed_genus")] <- "genus"
```

@@

```{r}
na_traits_after_genus <- filled_na_traits_family.genus %>%
  filter(is.na(TraitValue))
nrow(na_traits_after_genus)
filled_traits_with_genus <- filled_na_traits_family.genus %>%
  filter(!is.na(TraitValue))
nrow(filled_traits_with_genus)
```

@@

```{r}
filled_traits_with_family <- filled_na_traits_family %>%
  filter(!is.na(TraitValue))
nrow(filled_traits_with_family)
na_traits_after_family <- filled_na_traits_family %>%
  filter(is.na(TraitValue))
nrow(na_traits_after_family)
```

@@

```{r}
# Count number of NA records
na_records_count <- sum(is.na(all_traits_with_NAs$TraitValue))

# Count number of unique species with NA records
species_with_na_count <- all_traits_with_NAs %>%
  filter(is.na(TraitValue)) %>%
  distinct(species) %>%
  nrow()

# Count number of unique traits with NA records
traits_with_na_count <- all_traits_with_NAs %>%
  filter(is.na(TraitValue)) %>%
  distinct(TraitName) %>%
  nrow()

# Print results
cat("Number of NA records:", na_records_count, "\n")
cat("Number of species with NA records:", species_with_na_count, "\n")
cat("Number of traits with NA records:", traits_with_na_count, "\n")
```

@@

## Convert long dataframe to wide
```{r}
#remove unused columns
all_traits_with_NAs_long <- all_traits_with_NAs %>%
  select(-TraitLevel, -family, -genus) %>%
  distinct()
wide_traits <- pivot_wider(all_traits_with_NAs_long, names_from = TraitName, values_from = TraitValue)
```

```{r}
glimpse(wide_traits)
```

```{r}
#convert column types from character to factor & numeric
wide_traits$PlantLifespan_years <- as.numeric(wide_traits$PlantLifespan_years)
wide_traits$GrowthForm <- as.factor(wide_traits$GrowthForm)
wide_traits$SeedMass_g <- as.numeric(wide_traits$SeedMass_g)
wide_traits$PlantHeight_m <- as.numeric(wide_traits$PlantHeight_m)
wide_traits$FruitType <- as.factor(wide_traits$FruitType)
wide_traits$FruitLength_mm <- as.numeric(wide_traits$FruitLength_mm)
wide_traits$SeedWidth_mm <- as.numeric(wide_traits$SeedWidth_mm)
wide_traits$FruitMass_mg <- as.numeric(wide_traits$FruitMass_mg)
wide_traits$SeedLength_mm <- as.numeric(wide_traits$SeedLength_mm)
wide_traits$DispersalSyndrome <- as.factor(wide_traits$DispersalSyndrome)
```

@@

# Perform the imputation
```{r}
imp_model <- mice(wide_traits, method = "cart", maxit = 20)
```

@@

```{r}
glimpse(imputed_data)
```

# Summary
```{r}
fb_plot_species_traits_completeness(wide_traits)
ggsave("plant_trait_completeness_familygenus.png", plot = last_plot(), path = figure_path)
fb_plot_number_species_by_trait(wide_traits)
ggsave("plant_number_species_trait_familygenus.png", plot = last_plot(), path = figure_path)
fb_table_trait_summary(wide_traits)
```

```{r}
fb_plot_species_traits_completeness(imputed_data)
ggsave("plant_trait_completeness_imputed.png", plot = last_plot(), path = figure_path)
fb_plot_number_species_by_trait(imputed_data)
ggsave("plant_number_species_trait_imputed.png", plot = last_plot(), path = figure_path)
fb_table_trait_summary(imputed_data)
```

```{r}
# Reassign NA values in TraitLevel as "Imputed"
all_traits_with_NAs$TraitLevel[is.na(all_traits_with_NAs$TraitLevel)] <- "Imputed"

# Count the number of traits per TraitLevel and TraitName
trait_counts <- all_traits_with_NAs %>%
  count(TraitLevel, TraitName) %>%
  mutate(TraitLevel = fct_relevel(TraitLevel, "species", "genus", "family", "Imputed"))

# Create the faceted bar plot with free y-axis scales
ggplot(trait_counts, aes(x = TraitLevel, y = n, fill = TraitLevel)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Trait Counts per Level",
       x = "Trait Level",
       y = "Trait Count",
       fill = "Level") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none") +  # Remove legend
  facet_wrap(~ TraitName, scales = "free_y") +
  guides(fill = "none")  # Remove legend
ggsave("plant_trait_counts_per_level.png", plot = last_plot(), path = figure_path)
```

```{r}
# Count the number of traits per TraitLevel
trait_counts_overall <- all_traits_with_NAs %>%
  count(TraitLevel) %>%
  mutate(TraitLevel = fct_relevel(TraitLevel, "species", "genus", "family", "Imputed"))

# Create the faceted bar plot with free y-axis scales
ggplot(trait_counts_overall, aes(x = TraitLevel, y = n, fill = TraitLevel)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Trait Counts per Level",
       x = "Trait Level",
       y = "Trait Count",
       fill = "Level") +
  theme_minimal() +
  guides(fill = "none")  # Remove legend
ggsave("plant_trait_counts_per_level_overall.png", plot = last_plot(), path = figure_path)
```
